---
title: "IPCHEM Data Statistics"
output:
  flexdashboard::flex_dashboard:
    # logo: ./logo/eu.jpg
    vertical_layout: scroll
    orientation: rows
    css: custom.css
---

```{r setup, include=FALSE}
library(readr)
library(stringr)
library(tidyverse)
library(DT)
library(plotly)
library(flexdashboard)
library(leaflet)
library(maps)
library(lubridate)
source("fnutils.R")
```

Sidebar {.sidebar}
=======================================================================

<p style="font-size:13px">
This dashboard shows the IPCHEM data integration statistics, considering the chemicals, countries, media, sampling years. It includes only the publicly available data, while data sets with restricted access are not counted here.
</p>

<hr>

<!-- <style> -->
<!--         strong { -->
<!--             font-weight: bold; -->
<!--         } -->
<!--         navbar { -->
<!--             display: none; -->
<!--         } -->
<!-- </style> -->


<h5>Data layers</h5>

<p style="font-size:13px">
<b>Substance:</b> in IPCHEM is the analyte measured in a sample. This includes mostly classical chemicals, but includes also analytes such as Particulate Matter (PM 2.5 or PM10) or process-generated materials such as dust.
</p>

<p style="font-size:13px">
<b>Record:</b>
<ul>
<li style="font-size:13px"><u>single measurement:</u> individual measurement made at one point in time
</li>
<li style="font-size:13px"><u>aggregated data:</u> a set of one statistical values (e.g. mean, percentile, etc.), obtained by aggregating several individual measurements
</li>
</ul>
</p>

<p style="font-size:13px">
<b>Dataset:</b> a collection of data, published or curated by a single agent, and available for access or download in one or more representations
</p>

<p style="font-size:13px">
<b>Package:</b> a file containing a subset of the data records of a dataset grouped by chemical, media and period. Cumulative packages help downloading whole datasets, divided by period in the case of regular data collections (e.g. EFSAMOPER) 
</p>

<p style="font-size:13px">
<b>Media:</b> medium which was sampled and analysed in the measurement
</p>

<p style="font-size:13px">
<b>Module:</b> thematic area of the measurements
</p>

<hr>

<h5>Usage</h5>

<p style="font-size:13px"> 
<b>Plots:</b> click on any item from the legend to remove it from the plot. Advanced features using top right tooltips.
</p>

<p style="font-size:13px">
<b>Tables:</b> default sorted in descending order by number of records
</p>

<p  style="font-size:13px">
<b>Map:</b> hover to show the number of records by country. 
</p>

<hr>
<p style="font-size:13px">Report generated on: <b>`r format(Sys.Date(), "%m/%d/%Y")`</b></p>


```{r import_data, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}

# load main data

url = "https://ipchem.jrc.ec.europa.eu/public/IPCHEM_public_dataset_packages.csv"

list_of_packages = get_csv_url(url)
list_of_packages = chg_col_names(list_of_packages)

country_iso = readRDS("./data/country_iso")

wmap <- maps::map("world", fill = TRUE, plot = FALSE)
```


```{r global, echo=FALSE, warning=FALSE}

list_of_packages = list_of_packages %>% 
  dplyr::filter(!is.na(chemical_name))

# counts
n_datasets = length(unique(list_of_packages$dataset))
n_packages = length(unique(list_of_packages$package)) # NA is counted as 1. there are 2494 of them
n_chemicals = length(unique(list_of_packages$chemical_name[!is.na(list_of_packages$chemical_name)]))

# sums

sum_records = sum(list_of_packages$records)

# create pivot tables
pivot_media <- list_of_packages %>%
  group_by(media) %>%
  summarize(
    packages = n(),
    records = sum(records)
    ) %>% 
  ungroup() %>% 
  arrange(desc(records))

pivot_media_period <- list_of_packages %>%
  filter(!is.na(period)) %>% 
  group_by(media, period) %>%
  summarize(records = sum(records)) %>%
  mutate(records = cumsum(records))
  

# record and packages by dataset
pivot_dataset = list_of_packages %>% 
  group_by(dataset) %>% 
  summarize(
    packages = n(),
    records = sum(records)
    )%>% 
  ungroup() %>% 
  arrange(desc(records))

pivot_dataset_period = list_of_packages %>%
  filter(!is.na(period)) %>% 
  select(dataset, period, records) %>% 
  left_join(pivot_dataset %>%
              top_n(10, records) %>% 
              select(-packages, -records) %>% 
              mutate(top_10 = 1),
            by = join_by(dataset == dataset)) %>% 
  mutate(dataset = ifelse(is.na(top_10), 'OTHER', dataset)) %>% 
  group_by(dataset, period) %>% 
  summarize(records = sum(records)) %>% 
  mutate(records = cumsum(records))

# record by module and period
pivot_module = list_of_packages %>% 
  select(dataset, records) %>% 
  left_join(env_modules, by = join_by(dataset == dataset)) %>% 
  group_by(module) %>% 
  summarize(packages = n(),
    records = sum(records))%>% 
  ungroup() %>% 
  arrange(desc(records))

# record by module and period
pivot_module_period = list_of_packages %>% 
  filter(!is.na(period)) %>%
  select(dataset, period, records) %>% 
  left_join(env_modules, by = join_by(dataset == dataset)) %>% 
  group_by(module, period) %>% 
  summarize(records = sum(records)) %>%
  mutate(records = cumsum(records))

# split column location by each individual country and extract records
locations <- list_of_packages %>% 
  select(location, period, chemical_name) %>% 
  separate_rows(location, sep = ", ", convert = TRUE) %>% 
  mutate(country_code = str_split_fixed(location, ' \\[', 2)[, 1]) %>% 
  mutate(records = str_split_fixed(location, ' \\[', 2)[, 2]) %>%
  mutate(records = as.numeric(str_sub(records, 1, str_length(records) - 1))) %>% 
  select(-location)

# number or record by country
pivot_country = locations %>% 
  dplyr::filter(!is_null(country_code)) %>%
  dplyr::filter(country_code != "null") %>%
  group_by(country_code) %>% 
  summarize(records = sum(records))

pivot_region_period = locations %>% 
  inner_join(country_iso %>%
               select(country, country_code) %>% 
               distinct(), # some countries have more than 1 primary city
            by = join_by(country_code == country_code)) %>% 
  left_join(eu_27 %>% 
              mutate(region = 'EU27'),
            by = join_by(country_code == code)) %>% 
  mutate(region = ifelse(is.na(region), 'Outside_EU27', region)) %>% 
  group_by(region, period) %>%
  summarize(records = sum(records)) %>% 
  mutate(records = cumsum(records))

# which chemical has most of the records. top 10
chemical_pivot = list_of_packages %>% 
  filter(!is.na(chemical_name)) %>% 
  group_by(chemical_name) %>% 
  summarize(n = sum(records)) %>% 
  arrange(desc(n)) %>% 
  head(20)

# after split by country, sum the chemical records by country only for top chemicals
pivot_chemical_country = locations %>%
  inner_join(chemical_pivot %>% select(chemical_name),
             by = join_by(chemical_name == chemical_name)) %>%
  group_by(chemical_name, country_code) %>%
  summarize(records = sum(records))

# bring the nice name of the country
pivot_chemical_country = pivot_chemical_country %>% 
  left_join(country_iso %>% 
              select(country, country_code) %>% 
              distinct(), # some countries have more than 1 primary city
            by = join_by(country_code == country_code)) %>% 
  filter(!is.na(country)) %>% # 1A0 country code does not exist in coordinates. what is it?
  select(chemical_name, country, records)

# create map
# extract coordinates for each country capital 
country_coordinates <- pivot_country %>%
  left_join(country_iso, by = join_by(country_code == country_code)) %>%
  filter(!is.na(country))

# # remove duplicates
# country_coordinates = country_coordinates %>% 
#   group_by(country) %>% 
#   mutate(rnk = row_number()) %>% 
#   ungroup() %>% 
#   filter(rnk == 1) %>% 
#   select(-rnk)

# # normalize some names
# country_coordinates = country_coordinates %>%
#   mutate(country = ifelse(country == "Czechia", "Czech Republic", country)) %>%
#   mutate(country = ifelse(country == "Macedonia", "North Macedonia", country)) %>% 
#   mutate(country = ifelse(country == "United States", "USA", country)) %>% 
#   mutate(country = ifelse(country == "United Kingdom", "UK", country))

### the map
# color pallet based on records
pal = colorNumeric(palette = "YlGnBu", domain = country_coordinates$records, na.color = "#808080")
# pal = colorQuantile(palette = "YlGnBu", domain = country_coordinates$scores, n = 5, na.color = "#808080")


flag = stringr::str_locate(wmap$names, "\\w+:")
wmap$end = flag[, 2]
wmap$names = ifelse(is.na(wmap$end), wmap$names, str_sub(wmap$names, 1, wmap$end-1))


temp_df = data.frame(country = wmap$names)

temp_df = temp_df %>% 
   mutate(country = ifelse(country == "UK", "United Kingdom", country))

temp_df = temp_df %>% 
  left_join(country_coordinates %>% 
              select(country, records),
            by = join_by(country == country)) %>% 
  mutate(color = pal(records))

wmap$color = temp_df$color
wmap$records = temp_df$records

wmap$labels = paste(wmap$names, paste(":", ifelse(
  is.na(wmap$records), "", formatC(wmap$records, big.mark = ",", format = "f", digits = 0)
)), sep = "\n")

```

```{r}
# cleanup

rm(list_of_packages, locations, world_cities, temp_df)

```

Dashboard
=======================================================================

Row
-------------------------------------

### Number of Substances

```{r}

valueBox(prettyNum(n_chemicals, big.mark = ","), icon = "fa-database", color = " #7cabde")

```


### Number of Datasets

```{r}

valueBox(prettyNum(n_datasets, big.mark = ","), icon = "fa-database", color = " #7cabde")

```


### Number of Records

```{r}

valueBox(prettyNum(sum_records, big.mark = ","), icon = "fa-database", color = " #7cabde")

```


### Number of Packages

```{r}

valueBox(prettyNum(n_packages, big.mark = ","), icon = "fa-database", color = " #7cabde")

```



Row
-------------------------------------
### Top 20 substances recorded by country

```{r}

fig <- pivot_chemical_country %>% 
  mutate(color_code = as.character(as.integer(as.factor(country)))) %>% 
  mutate(country = str_sub(country, 1, 15)) %>% 
  plot_ly(
    x = ~ chemical_name,
    y = ~ records,
    type = 'bar',
    # name = ~ country,
    color = ~ country
  ) %>%
  layout(
    # title = 'Records per year',
    plot_bgcolor = '#e5ecf6',
    xaxis = list(
      title = 'Chemical',
      zerolinecolor = '#ffff',
      zerolinewidth = 2,
      gridcolor = 'ffff'
    ),
    yaxis = list(
      title = 'Number of Records',
      zerolinecolor = '#ffff',
      zerolinewidth = 2,
      gridcolor = 'ffff'
    )
  )

fig

```



Row
-------------------------------------

### Cumulative records by Module

```{r}

fig <-
  pivot_module_period %>%
  filter(period >= 1990) %>%
  mutate(symbol = as.factor(module)) %>%
  mutate(symbol = as.integer(symbol)) %>% 
  # mutate(module = str_sub(module, 1, 10)) %>% # so the legend won't be pushed right
  plot_ly(
    x = ~ period,
    y = ~ records,
    # data = pivot_country_period,
    type = 'scatter',
    mode = "markers",
    name = ~ module,
    symbol = ~ symbol,
    fill = 'tozeroy',
    marker = list(size = 10)
  ) %>%
  layout(
    # title = 'Records per year and country',
    plot_bgcolor = '#e5ecf6',
    xaxis = list(
      title = 'Year',
      zerolinecolor = '#ffff',
      zerolinewidth = 2,
      gridcolor = 'ffff'
    ),
    yaxis = list(
      title = 'Number of Records',
      zerolinecolor = '#ffff',
      zerolinewidth = 2,
      gridcolor = 'ffff'
    ),
    legend = list(x = 0.1, y = 0.9),
    showlegend = TRUE
  )

fig

```



### Cumulative records by Media
```{r}

fig <-
  pivot_media_period %>%
  filter(period >= 1990) %>%
  mutate(symbol = as.factor(media)) %>%
  mutate(symbol = as.integer(symbol)) %>% 
  # mutate(media = str_sub(media, 1, 10)) %>% # so the legend won't be pushed right
  plot_ly(
    x = ~ period,
    y = ~ records,
    # data = pivot_country_period,
    type = 'scatter',
    mode = "markers",
    name = ~ media,
    symbol = ~ symbol,
    fill = 'tozeroy',
    marker = list(size = 10)
  ) %>%
  layout(
    # title = 'Records per year and dataset',
    plot_bgcolor = '#e5ecf6',
    xaxis = list(
      title = 'Year',
      zerolinecolor = '#ffff',
      zerolinewidth = 2,
      gridcolor = 'ffff'
    ),
    yaxis = list(
      title = 'Number of Records',
      zerolinecolor = '#ffff',
      zerolinewidth = 2,
      gridcolor = 'ffff'
    ),
    legend = list(x = 0.1, y = 0.9),
    showlegend = TRUE
  )

fig

```

### Cumulative records by Dataset
```{r}

symbols <- c("circle", 'circle-open-dot', 'triangle-left-open', 'triangle-down-dot', 'cross-open', 'x-open', 'triangle-down', 'square', '4', '2')

fig <-
  pivot_dataset_period %>%
  filter(period >= 1990) %>% 
  mutate(symbol = as.factor(dataset)) %>%
  mutate(symbol = as.integer(symbol)) %>% 
  # mutate(records = log10(records)) %>% 
  # mutate(dataset = str_sub(dataset, 1, 10)) %>% # so the legend won't be pushed right
  plot_ly(
    x = ~ period,
    y = ~ records,
    # data = pivot_country_period,
    type = 'scatter',
    mode = "markers",
    name = ~ dataset,
    symbol = ~ symbol,
    symbols = symbols,
    fill = 'tozeroy',
    # alpha = 0.5, 
    # alpha_stroke = 1,
    marker = list(size = 10)
  ) %>%
  layout(
    # title = 'Records per year and dataset',
    plot_bgcolor = '#e5ecf6',
    xaxis = list(
      title = 'Year',
      zerolinecolor = '#ffff',
      zerolinewidth = 2,
      gridcolor = 'ffff'
    ),
    yaxis = list(
      title = 'Number of Records',
      zerolinecolor = '#ffff',
      zerolinewidth = 2,
      gridcolor = 'ffff'
    ),
    legend = list(x = 0.1, y = 0.9),
    showlegend = TRUE
  )

fig
```

Row
-------------------------------------

### Statistics by Module

```{r}

DT::datatable(pivot_module %>% 
                mutate(`%records` = 100 * records / sum(records)) %>%   
                mutate(packages = prettyNum(packages, big.mark = ",")) %>% 
                mutate(records = prettyNum(records, big.mark = ",")) %>% 
                mutate(`%records` =  formatC(`%records`, format = "f", digits = 2)),
              options = list(dom = 't',
                             bPaginate = FALSE,
                             columnDefs = list(list(className = 'dt-right', targets = 1:3))),
              rownames = FALSE)

```

### Statistics by Media

```{r}
DT::datatable(pivot_media %>% 
                mutate(`%records` = 100 * records / sum(records)) %>%   
                mutate(packages = prettyNum(packages, big.mark = ",")) %>% 
                mutate(records = prettyNum(records, big.mark = ",")) %>% 
                mutate(`%records` =  formatC(`%records`, format = "f", digits = 2)),
              options = list(dom = 't',
                             bPaginate = FALSE,
                             columnDefs = list(list(className = 'dt-right', targets = 1:3))),
              rownames = FALSE) 

```

### Statistics by Dataset

```{r}


DT::datatable(pivot_dataset %>% 
                mutate(`%records` = 100 * records / sum(records)) %>%   
                mutate(packages = prettyNum(packages, big.mark = ",")) %>% 
                mutate(records = prettyNum(records, big.mark = ",")) %>% 
                mutate(`%records` =  formatC(`%records`, format = "f", digits = 2)),
              options = list(dom = 't',
                             bPaginate = FALSE,
                             columnDefs = list(list(className = 'dt-right', targets = 1:3))),
              rownames = FALSE)

```




Row
-------------------------------------

### Map distribution of data records

```{r}


mp = leaflet(data = wmap,
             options = leafletOptions(
               minZoom = 2,
               maxZoom = 4,
               center = c(50.0875, 14.4214),
               zoom = 3
             )) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~ color,
    fillOpacity = 0.5,
    stroke = TRUE,
    color = "black",
    dashArray = "3",
    weight = 1,
    label = ~ labels,
    popup = ~ names
  ) %>%
  addLegend(
    "bottomright",
    pal = pal,
    values = ~ records,
    title = "Records",
    opacity = 1
  )

mp

```


### Cumulative country level records by year

```{r}

fig <-
  pivot_region_period %>%
  filter(period >= 1990) %>% 
  mutate(symbol = as.factor(region)) %>%
  mutate(symbol = as.integer(symbol)) %>%
  # mutate(country = str_sub(country, 1, 13)) %>% # so the legend won't be pushed right
  plot_ly(
    x = ~ period,
    y = ~ records,
    # data = pivot_country_period,
    type = 'scatter',
    mode = "lines+markers",
    name = ~ region,
    symbol = ~ symbol,
    fill = 'tozeroy',
    marker = list(size = 10)
  ) %>%
  layout(
    # title = 'Records per year and country',
    plot_bgcolor = '#e5ecf6',
    xaxis = list(
      title = 'Year',
      zerolinecolor = '#ffff',
      zerolinewidth = 2,
      gridcolor = 'ffff'
    ),
    yaxis = list(
      title = 'Number of Records',
      zerolinecolor = '#ffff',
      zerolinewidth = 2,
      gridcolor = 'ffff'
    ),
    legend = list(x = 0.1, y = 0.9),
    showlegend = TRUE
  )


fig

```

